Titeln på avhandlingens språk

### 2023-05-05 body/part1/chapter1-introduction/introduction.tex
images
fixes suggested by markus

\section{Introduction}

@@

\gls{XR}, on the other hand, is a broad term that encompasses various forms of immersive application systems.
\gls{XR} applications, such as \gls{AR}, \gls{VR}, or \gls{MR}, involve the real-time interaction between user and virtual elements in a partially to fully immersive environment.
For instance, \gls{MAR} is an application of \gls{AR} that involves overlaying digital information onto the physical world using a mobile device.
This definition has in later years been extended to encompass \gls{AR} deployed on wearable devices (such as Google Glass~\cite{googleglass}) as well.
\gls{MAR} allows users to, for example, view real-time information and details about products while in-store, receive instructions and guidance to perform a complex task, or contextual information while visiting a new city.
By using Edge Computing and 5G, \gls{MAR} --- and other \gls{XR} systems --- can operate with much greater responsiveness and reliability, enabling highly immersive and interactive experiences.

@@

The evaluation of performance in Edge computing systems has been addressed in existing literature through analytical, simulation, and experimental approaches.
However, these systems exhibit several characteristic that make their study complex.
One of these is unique nature of Edge computing environments, which tightly integrates compute and network.
This demands a wide range of interdisciplinary expertise to study, spanning (but not limited to) computer science, communication networks, information theory, queuing analysis, etc.
Such expertise may not always be feasible to obtain within a single research group.

@@

This has led to existing approaches to performance evaluation of Edge Computing environments to often be constrained in their scope, focusing on comprehensive characterization of either compute or network aspects, resulting in potential trade-offs that may neglect one half of the larger picture.
We argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
There is a need for holistic approaches that consider both network and compute characteristics of Edge workloads for accurate performance evaluation and optimization.

@@

\subsection{Structure of this dissertation}

This dissertation is structured into two parts.
\cref{part:summary} presents a summary of the research, introducing the topic, discussing related work, and highlighting key contributions.
Next, in \cref{part:publications} we present the publications that form the core of this thesis.

\cref{part:summary} is itself structured into four main chapters.
\cref{chap:introduction} serves as an introduction to the topic and outlines the key contributions of this work.
It sets the stage for the subsequent chapters by providing a high-level overview of the research problem and its significance, as well as provides background information on the core topics of the thesis.
In \cref{chap:relwork}, we present the scope of the thesis, as well as relevant related work.
The chapter provides a comprehensive review of the existing literature, highlighting the current state-of-the-art and identifying any gaps in knowledge.
We discuss context for our research, and establish the foundation for the contributions presented in the dissertation.
\cref{chap:contributions} is the core of \cref{part:summary} and provides a detailed summary of the contributions made in this dissertation.
We discuss methods, tools, and findings that were developed and obtained throughout the course of our research.
We aim to demonstrate the significance of our work and how it contributes to the field.
Finally, \cref{chap:conclusions} concludes the dissertation and outlines avenues for future research.
We summarize the key findings and contributions of the thesis, and briefly discuss the implications of our findings and contributions for the field.
This chapter also identifies open questions and areas for further research, providing opportunities to build on the work presented in this thesis.

>>>>

\section{Motivation}

@@

\gls{XR}, on the other hand, is a broad term that encompasses various forms of immersive application systems.
\gls{XR} applications, such as \gls{AR}, \gls{VR}, or \gls{MR}, involve the real-time interaction between user and virtual elements in a partially to fully immersive environment.
For instance, \gls{MAR} is an application of \gls{AR} that involves overlaying digital information onto the physical world using a mobile device.
This definition has in later years been extended to encompass \gls{AR} deployed on wearable devices (such as Google Glass~\cite{googleglass}) as well.
\gls{MAR} allows users to, for example, view real-time information and details about products while in-store, receive instructions and guidance to perform a complex task, or contextual information while visiting a new city.
An  example of such an application is illustrated in \cref{fig:augmentedreality}.
By using Edge Computing and 5G, \gls{MAR} --- and other \gls{XR} systems --- can operate with much greater responsiveness and reliability, enabling highly immersive and interactive experiences.

\begin{figure}
    \centering
    \includegraphics[height=26em]{figures/AR}
    \caption[{ }]{
        Example of an \gls{AR} application that overlays anatomical information on top of a live video feed of a model of human anatomy.\\

        {
            \footnotesize%
            Photo source: Pixabay~\cite{picture:AR}.
            Licensed under a \gls{CC0} license.
        }
    }\label{fig:augmentedreality}
\end{figure}

@@

The evaluation of performance in Edge computing systems has been addressed in existing literature through analytical~\cite{chen2015efficient,champati2015one,champati2016semi,al2017reliable}, simulation~\cite{svorobej2019simulating,gupta2017ifogsim,sonmez2018edgecloudsim,qayyum2018fognetsimpp}, and experimental~\cite{das2018edgebench,lee2019iotbench,george2020openrtist,mcchesney2019defog,baurle2022comb} approaches.
However, these systems exhibit several characteristic that make their study complex.
One of these is unique nature of Edge computing environments, which tightly integrates compute and network.
This demands a wide range of interdisciplinary expertise to study, spanning (but not limited to) computer science, communication networks, information theory, queuing analysis, etc.
Such expertise may not always be feasible to obtain within a single research group.

@@

This has led to existing approaches to performance evaluation of Edge Computing environments to often be constrained in their scope, focusing on comprehensive characterization of either compute or network aspects (e.g.~\cite{chen2015efficient} and~\cite{al2017reliable}), resulting in potential trade-offs that may neglect one half of the larger picture.
We argue that this limitation has led to an incomplete understanding of the overall performance of Edge systems and applications in the existing literature.
There is a need for holistic approaches that consider both network and compute characteristics of Edge workloads for accurate performance evaluation and optimization.

@@

### 2023-05-05 body/part1/chapter2-relatedwork/relatedwork.tex
contrib tree
images
data centers -> datacenters
fixes suggested by markus

\section{State of the art}\label{sec:state-of-the-art}

\subsection{Performance evaluation of Edge Computing systems}
\subsubsection{Analytical modeling and simulation approaches}

@@

Finally, another example is \emph{FogNetSim++}~\cite{qayyum2018fognetsimpp}, tool which builds upon the \acs{OMNET} network simulator.
FogNetSim++ allows users to simulate large Fog networks while incorporating mobility and scheduling models.
However, the tool does not include support for compute (either \gls{VM}- or container-based) migration within the network, a limitation which impacts the ability to accurately and simulate more complex scenarios in Fog and Edge Computing.

@@

% This dissertation complements existing work on benchmarking distributed system architectures.
% A rich body of literature exists with examples of such approaches for Cloud computing.
% For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of data centers and Cloud services; workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
% In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the Cloud.
% The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and includes generators that mimic remote clients accessing the system over the internet.\@
% \cite{malawski2018benchmarking,back2018using} focus instead on serverless computing, using compute-intensive workloads to benchmark major Cloud function providers such as \gls{AWS}, Microsoft Azure, and \gls{GCP}.

@@

\subsection{Experimental research in \glsfmtshort{MAR} and \glsfmtshort{CPS}}

>>>>

\section{State of the art}\label{sec:state-of-the-art}
\subsection{Performance evaluation of Edge Computing systems}
\subsubsection{Analytical modeling and simulation approaches}

@@

Finally, another example is \emph{FogNetSim++}~\cite{qayyum2018fognetsimpp}, a tool which builds upon the \acs{OMNET} network simulator.
FogNetSim++ allows users to simulate large Fog networks while incorporating mobility and scheduling models.
However, the tool does not include support for compute (either \gls{VM}- or container-based) migration within the network, a limitation which impacts the ability to accurately and simulate more complex scenarios in Fog and Edge Computing.

@@

% This dissertation complements existing work on benchmarking distributed system architectures.
% A rich body of literature exists with examples of such approaches for Cloud computing.
% For example, the \gls{YCSB}~\cite{cooper2010benchmarking} is a data-intensive benchmark for the profiling of datacenters and Cloud services; workload characterization in datacenters leads to a similar benchmark in~\cite{jia2013characterizing}.
% In~\cite{turner2012cmart}, the authors introduce \emph{C-MART}, a benchmark for modern web applications running on the Cloud.
% The benchmark utilizes modern web technologies, can be deployed automatically at a large scale, and includes generators that mimic remote clients accessing the system over the internet.\@
% \cite{malawski2018benchmarking,back2018using} focus instead on serverless computing, using compute-intensive workloads to benchmark major Cloud function providers such as \gls{AWS}, Microsoft Azure, and \gls{GCP}.

@@

\subsection{Experimental research in \glsfmtshort{MAR} and \glsfmtshortpl{NCS}}

### 2023-05-05 body/part1/chapter3-contributions/wca-implications.tex
fixes suggested by markus

On the other hand, in \cref{fig:ttfvsexectime:slice} we observe the evolution of execution times over \emph{slices} of \num{4} steps within blocks.
This figure illustrates how, at the lowest levels of impairment, users tended to get faster as the task progressed.
At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block was roughly \textasciitilde\SI{26}{\percent} lower than for steps \num{1} to \num{4}.
However, this effect is strongly dampened at higher levels of impairment, with the difference between the average execution times of steps in slices \num{2} and \num{3} basically disappearing at the highest \gls{TTF}.
This was also confirmed through an \gls{ANOVA} on factors slice and \gls{TTF}, finding significant (\ensuremath{p < 0.01}) for both factors and their interaction.

@@

Analyses of the correlation between individual subject differences and the above effects highlighted the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
This trait has been linked to low tolerance for stress, high emotional reactivity, as well as higher rates of \emph{delay discounting}~\cite{hirsh2008delay}.
Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \textasciitilde\SI{73}{\percent} of the variance in the execution time results, neuroticism was present in the first two.
Individual neuroticism scores were found to significantly linearly correlate (\ensuremath{rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.

@@

\subsubsection{Design of a realistic model for human timings in \gls{WCA}}

@@

In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:extendingmethodology}.
The discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.

>>>>

On the other hand, in \cref{fig:ttfvsexectime:slice} we observe the evolution of execution times over \emph{slices} of \num{4} steps within blocks.
This figure illustrates how, at the lowest levels of impairment, users tended to get faster as the task progressed.
At the lowest \gls{TTF}, the average execution time of steps \num{9} to \num{12} in a block was roughly \ensuremath{\sim\SI{26}{\percent}} lower than for steps \num{1} to \num{4}.
However, this effect is strongly dampened at higher levels of impairment, with the difference between the average execution times of steps in slices \num{2} and \num{3} basically disappearing at the highest \gls{TTF}.
This was also confirmed through an \gls{ANOVA} on factors slice and \gls{TTF}, finding significant (\ensuremath{p < 0.01}) for both factors and their interaction.

@@

Analyses of the correlation between individual subject differences and the above effects highlighted the \gls{BFI} trait of \emph{neuroticism} as an important modulating factor.
This trait has been linked to low tolerance for stress, high emotional reactivity, as well as higher rates of \emph{delay discounting}~\cite{hirsh2008delay}.
Out of the three components identified by a \gls{PCA}, which cumulatively accounted for \ensuremath{\sim\SI{73}{\percent}} of the variance in the execution time results, neuroticism was present in the first two.
Individual neuroticism scores were found to significantly linearly correlate (\ensuremath{\rho = 0.418}, two-tailed \ensuremath{p < 0.05}) with measured execution times at high \glspl{TTF}.

@@

\subsubsection{Design of a realistic model for human timings in \gls{WCA}}\label{sec:improved_model}

@@

In this section, we will discuss the implications of the methodology for the optimization of \gls{WCA} applications, in particular when employing the experimental results and the improved model of human timings discussed in \cref{sec:improved_model}.
The discussion presented here is a summarized version of what \cref{paper:olguinmunoz2023realistic} exposes.

### 2023-05-05 body/part1/chapter4-conclusions/conclusions.tex
clean up before sending to Markus
fixes suggested by markus

In this dissertation, we have presented two key contributions to the literature surrounding performance evaluation of Edge Computing systems.
Firstly, we have introduced a novel methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing, such as \gls{MAR} and \glspl{CPS}.
Our methodology is based on the key idea of emulating the end-user component of these applications while using real network and compute components at the backend, thus providing a comprehensive and realistic evaluation of system performance.
We have validated this methodology through two representative use-cases,\gls{WCA} and \glspl{NCS}.
Further, we have provided tools for its implementation, including two emulation toolkits for \gls{WCA} and \glspl{NCS}, as well as a software framework for the orchestration and automation of Edge Computing testbeds.
We have demonstrated the scalability and flexibility of our approach by deploying and studying scenarios with significant number of clients without the need for domain-specific hardware.

@@

Secondly, we have explored the implications of improved client-side emulation realism on the results produced with our methodology.
To this end, we have provided a novel characterization of human behavior in \gls{WCA}, which we used to design and implement an unprecedented model of human behavior for these applications.
We have then used this model to study the potential for optimization of these systems.
Our findings indicate that incorporating improved realism in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
Moreover, within the scope of \gls{WCA} --- and more broadly, \gls{MAR} --- our work emphasizes the importance of considering human behavior and reactions alongside system-related metrics or performance optimizations.
These behaviors play a critical role in the overall performance of such applications.

>>>>

In this dissertation, we have presented two key contributions to the literature surrounding performance evaluation of Edge Computing systems.
Firstly, we have introduced a novel methodology for the performance evaluation of latency-sensitive, feedback-loop-based applications enabled by Edge Computing, such as \gls{MAR} and \glspl{CPS}.
Our methodology is based on the key idea of emulating the end-user component of these applications while using real network and compute components at the backend, thus providing a comprehensive and realistic evaluation of system performance.
We have validated this methodology through two representative use-cases, \gls{WCA} and \glspl{NCS}.
Further, we have provided tools for its implementation, including two emulation toolkits for \gls{WCA} and \glspl{NCS}, as well as a software framework for the orchestration and automation of Edge Computing testbeds.
We have demonstrated the scalability and flexibility of our approach by deploying and studying scenarios with significant number of clients without the need for domain-specific hardware.

@@

Secondly, we have explored the implications of improved realism on the client-side emulation on the results produced with our methodology.
To this end, we have provided a novel characterization of human behavior in \gls{WCA}, which we used to design and implement an unprecedented model of human behavior for these applications.
We have then used this model to study the potential for optimization of these systems.
Our findings indicate that incorporating improved realism in client-side emulation can enable the implementation of otherwise unfeasible optimization approaches.
Moreover, within the scope of \gls{WCA} --- and more broadly, \gls{MAR} --- our work emphasizes the importance of considering human behavior and reactions alongside system-related metrics or performance optimizations.
These behaviors play a critical role in the overall performance of such applications.
